Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes

### 2020-05-28 Ben_Manuscripts/hdphmm/Draft.tex
Modified figures and edits for clarity

\title{Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series 
       of Solute Trajectories in Nanostructured Membranes}
\author{Benjamin J. Coscia}
\affiliation{Department of Chemical and Biological Engineering, University of Colorado Boulder, Boulder, CO 80309, USA}
\author{Christopher P. Calderon}
\affiliation{Ursa Analytics, Inc., Denver, CO 80212, USA} %CPC:  > 90% of my time spent officially under Ursa 
\alsoaffiliation{Department of Chemical and Biological Engineering, University of Colorado Boulder, Boulder, CO 80309, USA}
\author{Michael R. Shirts}
\email{michael.shirts@colorado.edu}
\affiliation{Department of Chemical and Biological Engineering, University of Colorado Boulder, Boulder, CO 80309, USA}

@@

  We can better understand the mechanisms controlling these dynamical modes by grouping 
  the states identified by the HDP-AR-HMM into clusters based on multiple metrics
  aimed at distinguishing solute behavior based on their fluctuations, dwell times
  in each state and position within the inhomogeneous membrane structure. We analyze
  prevalent clusters in order to relate their parameters to physical interactions between 
  solutes and the membrane.

@@

  There is a need for highly selective membranes in order to perform efficient 
  separations of the components of complex aqueous streams. Seawater and
  brackish water desalination, separation of organic micropollutants from 
  municipal water supplies, recovering valuable dissolved species from 
  waste streams and creating breathable barriers to harmful contaminants are just a
  subset of the broad range of applications for selective membrane separations.~\cite{fritzmann_state---art_2007,fonseca_couto_critical_2018,dischinger_evaluation_2019,ramaseshan_functionalized_2006}
  While researchers often focus on membrane permeability, many argue that
  it may be more beneficial to focus on membrane selectivity in order to 
  reduce the costs of commercial nanofiltration and reverse osmosis.~\cite{werber_materials_2016}

@@

  In this work we attempt uncover similar physical meaning through a data-driven process. 
  We utilize the sticky hierarchical Dirichlet process autoregressive hidden Markov model 
  (HDP-AR-HMM), a nonparameteric Bayesian classification technique that allows the user to 
  automatically detect and infer the parameters of an unknown number of latent vector
  autoregressive (VAR) modes present in solute center-of-mass MD 
  trajectory.~\cite{fox_bayesian_2010} The sticky HDP-AR-HMM greatly extends the 
  basic framework of simpler but more limited finite hidden Markov modeling approaches.

  In their most basic form, hidden Markov models (HMMs) are used to generate
  a series of latent, or hidden, states dependent only on the previous state.~\cite{rabiner_tutorial_1989}
  Propagating an HMM generates sequential but conditionally independent observations from 
  the emission distributions associated with each hidden state. 
  Within an HMM, dynamics (i.e. temporal dependence) are modeled solely through  
  a matrix of transition probabilities. Given a series of observations and a known number of hidden states, the 
  Baum-Welch algorithm, an expectation maximization technique, is a commonly 
  used way to determine the hidden state sequence and the underlying parameters of
  a standard HMM.~\cite{baum_maximization_1970} However, it is also possible to infer
  HMM parameters through a Bayesian framework.~\cite{scott_bayesian_2002,jasra_markov_2005} 
  One can also place priors on the transition probabilities and parameters of the hidden 
  states' emission distributions in order to iteratively infer their values and to
  estimate the hidden state sequence. HMMs have been used in a broad range of applications
  including speech recognition, weather forecasting and bioinformatics.~\cite{juang_hidden_1984,hughes_non-homogeneous_1999,yoon_hidden_2009}.
  In molecular simulations, HMMs have been used to characterize molecular kinetics
  in order to determine the equilibrium distribution of various observables.~\cite{thayer_hidden_2002,singhal_using_2004,noe_probability_2008,noe_projected_2013}

  By placing a hierarchical Dirichlet process (HDP) prior on the observed time
  series data, it is possible to extend HMMs to an unknown number of hidden
  states, thus allowing the model's complexity to be driven by the complexity
  of the data.~\cite{teh_hierarchical_2006} We will refer a model calibrated with
  this type of prior as an HDP-HMM; this model class is sometimes referred to
  as an infinite hidden Markov model.~\cite{beal_infinite_2002} The HDP-HMM 
  elegantly overcomes the main shortcoming of standard HMMs, the need to choose
  the number of states \textit{a priori}, making its use more advantageous in
  a diversity of applications including economic data and single particle trajectory 
  analysis.~\cite{shi_identifying_2016,hines_analyzing_2015}

  One issue with the original formulation of the HDP-HMM is that it does 
  not distinguish self-transitions and transitions between different states.
  Practically, this can result in rapid switching between states and inferred
  state sequence representations that do a poor job of describing the 
  observations.
  %BJC: added by Chris  
  Rapid state switching in the original HDP-HMM not only reduces the predictive
  ability of the model, but it is also physically unrealistic and can complicate
  model interpretation. 
  Fox et al. addressed this issue by adding a `sticky' 
  parameter.~\cite{fox_sticky_2007} The sticky HDP-HMM is useful in situations
  where the user has prior knowledge that state persistence is likely.
  For example, in the case of speaker diarization, it is unlikely that a 
  conversation between two people would rapidly switch between speakers.~\cite{fox_sticky_2011}
  This is also important for our work where we expect state persistence due
  to solute entrapment.

  The utility of the sticky HDP-HMM has been further extended to account for
  inherent temporal dependence between measurements common to many time series
  datasets. 
  %BJC: reworked by Chris
  For example, in both MD and single particle tracking, there is strong temporal
  dependence/correlation between successive measurements even within a 
  fixed state.~\cite{calderon_data-driven_2014}
  Fox et al. formulated a method that accounts for temporal dependence betwee
  successive observations by incorporating Markov jump linear systems (MJLSs)
  into infinite state models.~\cite{fox_nonparametric_2009} An MJLS models 
  each dynamical mode as a linear dynamical process. Two MJLS sub-classes are 
  switching linear dynamical systems (SLDSs) and switching vector autoregressive
  (VAR) processes. Switching VAR processes are a sub-class of SLDSs that do not
  incorporate measurement noise. Since MD simulations do not contain measurement
  noise, we will model the dynamical modes exhibited by MD solute trajectories 
  as VAR processes and we will refer to our model as an application of the 
  sticky HDP-AR-HMM in order to maintain its distinction from the even more
  general sticky HDP-SLDS models.

  % BJC: fair number of changes to this paragraph based on Chris' suggestions.
  % BJC: he really wants use to emphasize our contribution with this part
  To our knowledge, applying the HDP-AR-HMM in order to gain kinetic and
  mechanistic insight from MD simulation trajectories has not been done
  before. Typical applications of the HDP-AR-HMM center around determining the 
  number of states, the state sequence and analysis of the emission 
  distributions of a time series. The addition of temporal correlation to 
  the inference procedure has primarily been used to improve state sequence 
  identification.~\cite{calderon_inferring_2015} In our approach, we use the
  HDP-AR-HMM not only to segment a collection of MD time series into distinct 
  dynamical modes, but to uncover unknown transport mechanisms and create 
  improved predictive models. We cluster the parameters of individual states
  in order to reduce the total state space down to an interpretable size
  This allows us to learn the implied dominant physical mechanisms and describe
  solute kinetics based on interpretation of distinct parameters that dominate
  the ensemble of MD trajectories. We can use the model parameters determined 
  either before or after clustering to generate new time series trajectories 
  which enable projections of the MD-generated time series to arbitrary length
  time scales.

  We aim to address two primary scientific questions with this work:
  \begin{enumerate}
  \item First, we want to know whether it is possible to use the HDP-AR-HMM to 
  help us predict macroscopic transport properties. We use the VAR parameters
  and state transition probability matrix of the HDP-AR-HMM in order to 
  generate stochastic trajectory realizations with behavior similar to that
  exhibited by MD. One can project these realizations onto much longer 
  time scales with computational ease extending into time scales where 
  regular diffusive behavior again occurs.
  \item Second, we want to know if the HDP-AR-HMM algorithm can help
  efficiently uncover underlying transport mechanisms which give rise to different 
  solute dynamical behavior.
  We use the parameters of the states identified by the HDP-AR-HMM in order to infer 
  dominant solute-membrane interactions and transport mechanisms. We cluster similar 
  state parameters in order to reduce the state space and to understand which segments 
  of the solute trajectories exhibit similar behavior. We support our mechanistic 
  hypotheses via quantitative comparison to various physical membrane properties and
  solute-membrane interactions.
  \end{enumerate}

@@

  We ran all MD simulations and energy minimizations using GROMACS 2018.~\cite{bekker_gromacs:_1993,berendsen_gromacs:_1995,van_der_spoel_gromacs:_2005,hess_gromacs_2008}  
  We implemented the sticky HDP-AR-HMM in Python 
  (\texttt{https://github.com/bencoscia/hdphmm}) which we heavily adapted from
  the MATLAB code of Fox et al.~\cite{fox_nonparametric_2009} Since we do not incorporate
  any novel methodology to the HDP-AR-HMM inference procedure, we refer the 
  interested reader to an excellent and in-depth description of the formulation of
  the HDP-AR-HMM and related models.~\cite{fox_bayesian_2010} 
  %MRS: possibly should give the permanent reference to a shirtsgroup fork?  We can discuss which is likely to be more permanent  
  All other post-simulation trajectory analysis tools are available online at
  \texttt{https://github.com/shirtsgroup/LLC\_Membranes}.

@@

  We studied transport of solutes in the H\textsubscript{II} phase using an
  atomistic molecular model of four pores in a monoclinic unit cell with 
  10\% water by weight. Approximately one third of the water molecules 
  occupy the tail region with the rest near the pore center.~\cite{coscia_chemically_2019}

  We chose to study a subset of 4 of the fastest moving solutes from our previous
  work: methanol, acetic acid, urea and ethylene glycol. In addition to exploring 
  membrane structural space the most, these solutes have a relatively diverse set
  of chemical functionality. For each solute we created a separate system and to 
  each system we added 6 solutes per pore for a total of 24 solutes. This number 
  of solutes per pore provides a sufficient amount of data from which to generate 
  statistics on the time scales which we simulate and prevents solute clustering
  facilitating our study of pure solute-membrane interactions. Further details on
  the setup and equilibration of these systems are detailed in our previous work.~\cite{coscia_chemically_2019}

@@

  \textit{State sequence and parameter inference}: We ran 2000 iterations of the
  HDP-AR-HMM procedure on each trajectory which was sufficient in order to arrive
  at converged state sequences and VAR parameter estimates. By our criteria, the 
  procedure generally converged within 500--1000 iterations. We evaluated 
  convergence based on when the VAR parameter estimates plateaued. We detected 
  plateaus using the module \texttt{pymbar.timeseries.detect\_equilibration}~\cite{chodera_simple_2016} 
  on the time series of the individual parameters (Section~\ref{S-section:convergence}
  of the Supporting Information). Convergence of the VAR parameters implies that
  the state sequence is relatively stable since the inferred parameters are 
  sensitive to the inputs to the MNIW conjugate priors. The precise boundaries of 
  the state segments can fluctuate; therefore, in states with a low number of 
  emissions, the variance in $A$ and $\Sigma$ parameters can be high. In these 
  cases, plateauing of the mean in each dimension tends to be a more reliable 
  indicator of convergence (see Figure~\ref{S-fig:convergence3d} of the Supporting
  Information).

  It is usually beneficial to employ intelligent state sequence initialization procedures
  in nonparametric Bayesian approaches.~\cite{fox_bayesian_2010}
  Most importantly in this work, the states identified by the HDP-AR-HMM are heavily 
  influenced by the Gaussian prior placed on $\mathbf{c}$ in Equation~\ref{eqn:var}. In 
  Section~\ref{S-section:fitting_tips} of the Supporting Information we outline our 
  algorithmic method for choosing reliable prior parameters in addition to other 
  practical tips for initialization.

  %BJC: The following paragraph used to be the first paragraph after italics above. 
  % I moved it down here as suggested by Chris below. What do you think in terms of flow?  
  We inferred the state sequence and parameters for each of the 24 solute 
  trajectories independently. Although the HDP-AR-HMM is capable of identifying 
  an infinite number of states, a Dirichlet process tends to exhibit a 
  ``rich get richer" effect, favoring a fewer number of states.~\cite{dreyer_discovering_2011}
  By applying the algorithm to each trajectory independently, we reduce the 
  possibility of lumping together multiple similar states which we would prefer
  to stay separated before clustering. 
  %BJC: added following sentence to aid reader in response to Chris' question below 
  In Section~\ref{method:realizations}, we 
  explain how we use these 24 independent models in order to generate a single
  MSD prediction.
  % BJC: abridged Chris comment with respect to this paragraph here:
  % CPC: the question i have at this point as a reader in this area, given 
  % how this is setup both as predictive and clustering, is "how do they plan to
  % generate data and predict with this model if they have 24 different transition
  % mat estimates"
  % CPC: returning to comment after reading.  section 2.4 explains what you do 
  % (though i'm still fuzzy on how you estimate transition matrix probs in method 2). 
  % (BJC: Note I tried to clarify this process for method 2 below)
  % CPC_continued: in any case, you might want to avoid obvious critiques and head 
  % scratching from expert reviewers. i'd suggest moving this paragraph to end of subsection and
  % expand on above considerations after you've given more cut and dry details below and reference 2.4 stating     
  %that you say how you simulate from a model estimated one trajectory at a time via HDP-AR-HMM inference

@@

  We clustered states with like parameter sets in order to reduce the state space to
  a more easily interpretable size 
  %BJC: added following in response to Chris' below comment. I think I'm interpretting what he's saying right
  % CPC: perhaps keep hammering home not only reducing, but also reemphasize why absolute
  % value of some parameters really not of interest 
  and because the parameters of a single states does not always give meaningful physical insight. 
  For each solute studied, we identified approximately 300-500 independent states, 
  each with separate VAR(1) parameters. Many of these states exhibit very similar
  dynamical behavior except their mean levels in each dimension are different.

@@

  We clustered based on the eigenvalues of $A$ and $\Sigma$, the distance from the 
  pore center and the self-transition probabilities, $P_{ii}$, of each state. $A$ 
  and $\Sigma$ quantify the temporal correlation and size of fluctuations in a 
  given state. $P_{ii}$ is directly related to the expected time spent in each state.
  The distance from the closest pore center is expected to be physically meaningful
  since the membrane is radially inhomogeneous from each pore, transitioning from 
  hydrophilic pores to hydrophobic tails. However, the channels are isotropic along
  the pore axis so we do not cluster in means in the $z$ dimension. Before carrying
  out the agglomerative clustering algorithm, we normalized each clustering dimension
  by its variance and shifted the mean to zero.

  %BJC: small modifications to this paragraph for clarity
  We converted the parameters that are directly related to solute position into 
  cylindrical coordinates and cast them in the same units before clustering. 
  We replaced the $x$ and $y$ eigenvalues of $\Sigma$ and $A$ with $\lambda_x$ + 
  $\lambda_y$ and $\lambda_x^2$ + $\lambda_y^2$ respectively, which are invariant with 
  rotation around the $z$ axis. We clustered on the squared means in $r$, $\mu_r^2 = 
  \mu_x^2 + \mu_y^2$, where $\mu_r$, $\mu_x$ and $\mu_y$ are the means in $r$, $x$ and
  $y$ respectively.
%  rather than $\mu_x$ and $\mu_y$ individually, with the
%  square $\mu_r$ converting it into the same units as $A$ and $\Sigma$.

  To cluster on $P_{ii}$, we first cast it in terms of the expected dwell times. The expected
  value of the number of sequential self-transitions is simply:
  \begin{equation}  
   E[dwell~time] = \frac{1}{1 - P_{ii}}
   \label{eqn:dwell_times}
  \end{equation}  
  This relationship implies that dwell times approach infinity as $P_{ii}$ approaches
  1. Therefore we found the most success with agglomerative clustering by linearizing this
  relationship and clustering on $-\log(1 - P_{ii})$.

@@

  We obtained estimates of clustered $A$ and $\Sigma$ parameters as well as a
  single $T$, describing transitions between all possible clustered states exhibited by
  the ensemble of trajectories, using the inference component of the HDP-AR-HMM
  algorithm. We applied the HDP-AR-HMM simultaneously on all 24 zeroed solute
  trajectories with state sequences fixed according to the remapped clustered 
  state assignments. This is only possible because the state sequence is fixed. 
  In cases where one does not know the state sequence and would like to 
  incorporate data from multiple trajectories, one can apply the beta process (BP)
  AR-HMM.~\cite{fox_sharing_2009} However the data requirements may be quite high
  for a good parameterization, which is why we elected to follow the clustering 
  approach described thus far.
  %BJC: I actually found 3 different ways do this. All giving the same results.
  % Method 1: described above - this is the most efficient way
  % Method 2: Link all 24 trajectories by adding a coffin state (basically a 
  % state that is a vector of zeros) between each traj
  % Method 3: Parameterize each trajectory independently, average together. T is
  % estimated by making a count matrix based on inputs from each trajectory

@@

  For both approaches, we drew state sequences with transition probabilities given 
  % BJC: Chris wasn't clear where T came from for method 2. I tried to make it clearer in 
  % last paragraph of previous section. Let me know if it's not
  by $T$ of the HDP-AR-HMM. While in a given state, we simulated motion according to the 
  VAR(1) parameterization of that state. After each state transition, we set the 
  unconditional mean of the new state to the particle's position immediately before
  the state transition occurred.

@@

  For the MD MSDs, we bootstrapped according to Method 1, sampling with 
  replacement from the ensemble of 24 solute trajectories.

@@

  We applied the HDP-AR-HMM independently to all 24 trajectories of each of the four solutes studied.
  The number of states found for each trajectory varied between 2 and 45.  %BJC4: MET: (5, 28), URE: (5, 45), ACH: (2, 25), GCL: (7, 35)
  Figure~\ref{fig:rz_unclustered} shows the state sequence determined by the HDP-AR-HMM for
  an example methanol trajectory. The model distinguishes many of the states due to 
  differences in their mean levels while their fluctuations about those means are often 
  similar. In Section~\ref{section:mechanisms}, we use clustering to reduce the total state
  space for each solute in order to more easily interpret the state parameters.

@@

  %BJC6: added. Should I add a figure?
  A potential shortcoming of the HDP-AR-HMM applied to MD simulations is its 
  inability to reproduce the short time scale MSD curves. The MSDs predicted
  by the model tend to enter a linear regime much more quickly than the MD MSDs
  (see Figure~\ref{S-fig:unclustered_msd_shortlag} of the Supporting Information
  for example). This is not a major concern for predicting macroscopic properties
  since we are primarily interested in reproducing the MSDs on much longer 
  timescales.
  %BJC6: could hypothesize why using previous work as support

>>>>

\title{Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series 
       of Solute Trajectories in Nanostructured Membranes}
\author{Benjamin J. Coscia}
\affiliation{Department of Chemical and Biological Engineering, University of Colorado Boulder, Boulder, CO 80309, USA}
\author{Christopher P. Calderon}
\affiliation{Ursa Analytics, Inc., Denver, CO 80212, USA}
\alsoaffiliation{Department of Chemical and Biological Engineering, University of Colorado Boulder, Boulder, CO 80309, USA}
\author{Michael R. Shirts}
\email{michael.shirts@colorado.edu}
\affiliation{Department of Chemical and Biological Engineering, University of Colorado Boulder, Boulder, CO 80309, USA}

@@

  We can better understand the mechanisms controlling these dynamical modes by grouping 
  the states identified by the HDP-AR-HMM into clusters based on multiple metrics
  aimed at distinguishing solute behavior based on their fluctuations, dwell times
  in each state and position within the inhomogeneous membrane structure. We analyze
  predominate clusters in order to relate their dynamical parameters to physical
  interactions between solutes and the membrane.

@@

  There is a need for highly selective membranes in order to perform efficient 
  separations of the components of complex aqueous streams. Seawater and
  brackish water desalination, separation of organic micropollutants from 
  municipal water supplies, recovery of valuable dissolved species from 
  waste streams and the creation of breathable barriers to harmful contaminants are just a
  subset of the broad range of applications for selective membrane separations.~\cite{fritzmann_state---art_2007,fonseca_couto_critical_2018,dischinger_evaluation_2019,ramaseshan_functionalized_2006}
  While researchers often focus on membrane permeability, many argue that
  it may be more beneficial to focus on membrane selectivity in order to 
  reduce the costs of commercial nanofiltration and reverse osmosis.~\cite{werber_materials_2016}

@@

  In this work we attempt to uncover similar physical meaning through a data-driven process. 
  We utilize the sticky hierarchical Dirichlet process autoregressive hidden Markov model 
  (HDP-AR-HMM), a nonparameteric Bayesian classification technique that allows the user to 
  automatically detect and infer the parameters of an unknown number of latent vector
  autoregressive (VAR) modes present in solute center-of-mass MD 
  trajectories.~\cite{fox_bayesian_2010} The sticky HDP-AR-HMM greatly extends the 
  basic framework of simpler but more limited finite hidden Markov modeling approaches.

  In their most basic form, hidden Markov models (HMMs) are used to generate
  a series of latent, or hidden, states dependent only on the previous state.~\cite{rabiner_tutorial_1989}
  Propagating an HMM generates sequential but conditionally independent observations from 
  the emission distributions associated with each hidden state. 
  Within an HMM, dynamics (i.e. temporal dependence) are modeled solely through  
  a matrix of transition probabilities. Given a series of observations and a known number of hidden states, the 
  Baum-Welch algorithm, an expectation maximization technique, is a commonly 
  used way to determine the hidden state sequence and the underlying parameters of
  a standard HMM.~\cite{baum_maximization_1970} However, it is also possible to infer
  HMM parameters through a Bayesian framework.~\cite{scott_bayesian_2002,jasra_markov_2005} 
  One can place priors on the transition probabilities and parameters of the hidden 
  states' emission distributions in order to iteratively infer their values and to
  estimate the hidden state sequence. HMMs have been used in a broad range of applications
  including speech recognition, weather forecasting and bioinformatics.~\cite{juang_hidden_1984,hughes_non-homogeneous_1999,yoon_hidden_2009}.
  In molecular simulations, HMMs have been used to characterize molecular kinetics
  in order to determine the equilibrium distribution of various observables.~\cite{thayer_hidden_2002,singhal_using_2004,noe_probability_2008,noe_projected_2013}

  By placing a hierarchical Dirichlet process (HDP) prior on the observed time
  series data, it is possible to extend HMMs to an unknown number of hidden
  states, thus allowing the model's complexity to be driven by the complexity
  of the data.~\cite{teh_hierarchical_2006} We will refer a model calibrated with
  this type of prior as an HDP-HMM; this model class is sometimes referred to
  as an infinite hidden Markov model.~\cite{beal_infinite_2002} The HDP-HMM 
  elegantly overcomes the main shortcoming of standard HMMs, the need to choose
  the number of states \textit{a priori}, making its use more advantageous in
  a diversity of applications including economic data and single particle trajectory 
  analysis.~\cite{shi_identifying_2016,hines_analyzing_2015,sgouralis_introduction_2017,jazani_alternative_2019,wang_single-molecule_2014,persson_extracting_2013}

  One issue with the original formulation of the HDP-HMM is that it does 
  not distinguish self-transitions and transitions between different states.
  Practically, this can result in rapid switching between states and inferred
  state sequence representations that do a poor job of describing the 
  observations. Rapid state switching in the original HDP-HMM not only 
  reduces the predictive ability of the model, but in many cases is also 
  physically unrealistic and can complicate model interpretation. 
  Fox et al. addressed this issue by adding a `sticky'
  parameter.~\cite{fox_sticky_2007} The sticky HDP-HMM is useful in situations
  where the user has prior knowledge that state persistence is likely.
  For example, in the case of speaker diarization, it is unlikely that a 
  conversation between two people would rapidly switch between speakers.~\cite{fox_sticky_2011}
  This is also important for our work where we expect state persistence due
  to solute entrapment.

  The utility of the sticky HDP-HMM has been further extended to account for
  inherent temporal dependence between measurements common to many time series
  datasets. For example, in both MD and single particle tracking, there is 
  strong temporal dependence/correlation between successive measurements even
  within a fixed state.~\cite{calderon_data-driven_2014}
  Fox et al. formulated a method that accounts for temporal dependence betwee
  successive observations by incorporating Markov jump linear systems (MJLSs)
  into infinite state models.~\cite{fox_nonparametric_2009} An MJLS models 
  each dynamical mode as a linear dynamical process. Two MJLS sub-classes are 
  switching linear dynamical systems (SLDSs) and switching vector autoregressive
  (VAR) processes. Switching VAR processes are a sub-class of SLDSs that do not
  incorporate measurement noise. Since MD simulations do not contain measurement
  noise, we will model the dynamical modes exhibited by MD solute trajectories 
  as VAR processes and we will refer to our model as an application of the 
  sticky HDP-AR-HMM in order to maintain its distinction from the even more
  general sticky HDP-SLDS models.

  To our knowledge, applying the HDP-AR-HMM in order to gain kinetic and
  mechanistic insight from MD simulation trajectories has not been done
  before. Typical applications of the HDP-AR-HMM center around determining the 
  number of states, the state sequence and analysis of the emission 
  distributions of a time series. The addition of temporal correlation to 
  the inference procedure has primarily been used to improve state sequence 
  identification.~\cite{calderon_inferring_2015} In our approach, we use the
  HDP-AR-HMM not only to segment a collection of MD time series into distinct 
  dynamical modes, but to uncover unknown transport mechanisms and to create 
  improved predictive models. We cluster the parameters of individual states
  in order to reduce the total state space down to an interpretable size
  This allows us to learn the implied dominant physical mechanisms and describe
  solute kinetics based on interpretation of distinct parameters that dominate
  the ensemble of MD trajectories. We can use the model parameters determined 
  either before or after clustering to generate new time series trajectories 
  which enable projections of the MD-generated time series to arbitrary length
  time scales.

  % BJC: Michael prefers not to put any conclusions in the intro. So we basically pose
  % questions to which we already know the answer, with the intent of readers having
  % a better sense of what to look for while reading.
  We aim to address two primary scientific questions with this work:
  \begin{enumerate}

  \item First, we want to know whether it is possible to use the HDP-AR-HMM to 
  help us predict macroscopic transport properties. We use the VAR parameters
  and state transition probability matrix of the HDP-AR-HMM in order to 
  generate stochastic trajectory realizations with behavior similar to that
  exhibited by MD. One can project these realizations onto much longer 
  time scales with computational ease extending into time scales where 
  regular diffusive behavior again occurs.

  \item Second, we want to know if the HDP-AR-HMM algorithm can help
  efficiently uncover underlying transport mechanisms which give rise to different 
  solute dynamical behavior. We cluster similar state parameters across all
  trajectories in order to reduce the total state space and to understand which
  segments of the solute trajectories exhibit similar behavior. We use the 
  parameters of the clustered states identified by the HDP-AR-HMM in order to 
  infer dominant solute-membrane interactions and transport mechanisms. We 
  support our mechanistic hypotheses via quantitative comparison to various 
  physical membrane properties and solute-membrane interactions.

  \end{enumerate}

@@

  We ran all MD simulations and energy minimizations using GROMACS 2018.~\cite{bekker_gromacs:_1993,berendsen_gromacs:_1995,van_der_spoel_gromacs:_2005,hess_gromacs_2008}  
  We implemented the sticky HDP-AR-HMM in Python 
  (\texttt{https://github.com/shirtsgroup/hdphmm}) which we heavily adapted from
  the MATLAB code of Fox et al.~\cite{fox_nonparametric_2009} Since we do not incorporate
  any novel methodology to the HDP-AR-HMM inference procedure, we refer the 
  interested reader to an excellent and in-depth description of the formulation of
  the HDP-AR-HMM and related models.~\cite{fox_bayesian_2010} 
  All other post-simulation trajectory analysis tools are available online at
  \texttt{https://github.com/shirtsgroup/LLC\_Membranes}.

@@

  We studied transport of solutes in the H\textsubscript{II} phase using an
  atomistic molecular model of four pores in a monoclinic unit cell with 
  10\% water by weight. Approximately one third of the water molecules 
  occupy the tail region with the rest are near the pore center.~\cite{coscia_chemically_2019}

  We chose to study a subset of 4 of the fastest moving solutes from our previous
  work: methanol, acetic acid, urea and ethylene glycol. In addition to exploring 
  membrane structural space the most, these solutes have a relatively diverse set
  of chemical functionality. For each solute we created a separate system and to 
  each system we added 6 solutes per pore for a total of 24 solutes. This number 
  of solutes per pore provides a sufficient amount of data from which to generate 
  statistics on the time scales which we simulate and prevents any significant
  solute clustering facilitating our study of pure solute-membrane interactions. 
  Further details on the setup and equilibration of these systems are detailed 
  in our previous work.~\cite{coscia_chemically_2019}

@@

  \textit{State sequence and parameter inference}: We ran 2000 iterations of the
  HDP-AR-HMM procedure on each trajectory which was sufficient in order to arrive
  at converged state sequences and VAR parameter estimates. By our criteria, the 
  procedure generally converged within 500--1000 iterations. We evaluated 
  convergence based on when the VAR parameter estimates plateaued. We detected 
  plateaus using the module \\ \texttt{pymbar.timeseries.detect\_equilibration}~\cite{chodera_simple_2016} 
  on the time series of the individual parameters (see Section~\ref{S-section:convergence}
  of the Supporting Information). Convergence of the VAR parameters implies that
  the state sequence is relatively stable since the inferred parameters are 
  sensitive to the inputs to the MNIW conjugate priors. The precise boundaries of 
  the state segments can fluctuate; therefore, in states with a low number of 
  emissions, the variance in $A$ and $\Sigma$ parameters can be high. In these 
  cases, plateauing of the mean in each dimension tends to be a more reliable 
  indicator of convergence (see Figure~\ref{S-fig:convergence3d} of the Supporting
  Information).

  It is usually beneficial to employ well-informed state sequence initialization 
  procedures in nonparametric Bayesian approaches.~\cite{fox_bayesian_2010}
  Most importantly in this work, the states identified by the HDP-AR-HMM are heavily 
  influenced by the Gaussian prior placed on $\mathbf{c}$ in Equation~\ref{eqn:var}. In 
  Section~\ref{S-section:fitting_tips} of the Supporting Information we outline our 
  algorithmic method for choosing reliable prior parameters in addition to other 
  practical tips for initialization.

  We inferred the state sequence and parameters for each of the 24 solute 
  trajectories independently. Although the HDP-AR-HMM is capable of identifying 
  an infinite number of states, a Dirichlet process exhibits a ``rich get richer" 
  effect, favoring a fewer number of states.~\cite{dreyer_discovering_2011}
  By applying the algorithm to each trajectory independently, we reduce the 
  possibility of lumping together multiple similar, but distinct states. In 
  Section~\ref{method:realizations}, we explain how we use these 24 independent
  models in order to generate a single MSD prediction.

@@

  We clustered states with like parameter sets in order to reduce the state space to
  a more easily interpretable size 
  % CPC: perhaps keep hammering home not only reducing, but also reemphasize why absolute
  % value of some parameters really not of interest 
  %MRS: explain why parameters of single states don't give insight; more context needed?  Is CPC talking about
  %MRS: some some of the parmameters not being interesting in general, not just the values for some states?
  and because the parameters of a single state alone does not always have enough context
  to provide meaningful physical insight. For each solute studied, we identified 
  approximately 300-500 independent states, each with separate VAR(1) parameters.
  Many of these states exhibit very similar dynamical behavior except their mean 
  levels in each dimension are different.

@@

  We clustered based on the eigenvalues of $A$ and $\Sigma$, the distance from the 
  pore center and the self-transition probabilities, $P_{ii}$, of each state. $A$ 
  and $\Sigma$ quantify the temporal correlation and size of fluctuations in a 
  given state. $P_{ii}$ is directly related to the expected time spent in each state.
  The distance from the closest pore center is expected to be physically meaningful
  since the membrane is radially inhomogeneous from each pore, transitioning from 
  hydrophilic pores to hydrophobic tails. However, the channels are isotropic along
  the pore axis so we do not cluster on the means in the $z$ dimension. Before carrying
  out the agglomerative clustering algorithm, we normalized each clustering dimension
  by its variance and shifted its mean to zero.

  We converted the parameters that are directly related to solute position into 
  cylindrical coordinates and cast them in the same units before clustering. 
  We replaced the $x$ and $y$ eigenvalues of $\Sigma$ and $A$ with $\lambda_x$ + 
  $\lambda_y$ and $\lambda_x^2$ + $\lambda_y^2$ respectively, which are invariant with 
  rotation around the $z$ axis. We clustered on the squared means in $r$, $\mu_r^2 = 
  \mu_x^2 + \mu_y^2$, where $\mu_r$, $\mu_x$ and $\mu_y$ are the means in $r$, $x$ and
  $y$ respectively.

  To cluster on $P_{ii}$, we first cast them in terms of the expected dwell times. The expected
  value of the number of sequential self-transitions is simply:
  \begin{equation}  
   E[dwell~time] = \frac{1}{1 - P_{ii}}
   \label{eqn:dwell_times}
  \end{equation}  
  This relationship implies that dwell times approach infinity as $P_{ii}$ approaches
  1. Therefore we found the most success with agglomerative clustering by linearizing this
  relationship and clustering on $-\log(1 - P_{ii})$.

@@

  We obtained estimates of clustered $A$ and $\Sigma$ parameters as well as a
  consensus $T$ matrices, describing transitions between all possible clustered states 
  exhibited by the ensemble of trajectories, using the inference component of 
  the HDP-AR-HMM algorithm. We applied the HDP-AR-HMM simultaneously to all 24
  zeroed solute trajectories with state sequences fixed according to the remapped 
  clustered state assignments. Simultaneous application of the HDP-AR-HMM to 
  all trajectories as we have done is only possible because the state sequence 
  is fixed. In cases where one does not know the state sequence and would like to 
  incorporate data from multiple trajectories, one can apply the beta process (BP)
  AR-HMM.~\cite{fox_sharing_2009} However the data requirements may be quite high
  for a good parameterization, which is why we elected to follow the clustering 
  approach described thus far.
  %BJC: I actually found 3 different ways do this. All giving the same results.
  % Method 1: described above - this is the most efficient way
  % Method 2: Link all 24 trajectories by adding a coffin state (basically a 
  % state that is a vector of zeros) between each traj
  % Method 3: Parameterize each trajectory independently, average together. T is
  % estimated by making a count matrix based on inputs from each trajectory

@@

  For both approaches, we drew state sequences with transition probabilities given 
  by $T$ of the HDP-AR-HMM. While in a given state, we simulated motion according to the 
  VAR(1) parameterization of that state. After each state transition, we set the 
  unconditional mean of the new state to the particle's position immediately before
  the state transition occurred.

@@

  For the MD MSDs, we bootstrapped according to Method 1, sampling with 
  replacement from the single ensemble of 24 solute trajectories.

@@

  We applied the HDP-AR-HMM independently to all 24 trajectories of each of the four solutes studied.
  The number of states found for each trajectory varied between 2 and 45.
  Figure~\ref{fig:rz_unclustered} shows the state sequence determined by the HDP-AR-HMM for
  an example methanol trajectory. The model distinguishes many of the states due to 
  differences in their mean levels while their fluctuations about those means are often 
  similar. In Section~\ref{section:mechanisms}, we use clustering to reduce the total state
  space for each solute in order to more easily interpret the state parameters.

@@

  A potential shortcoming of the HDP-AR-HMM applied to MD simulations is its 
  inability to reproduce the short time scale MSD curves. The MSDs predicted
  by the model tend to enter a linear regime much more quickly than the MD MSDs
  (see Figure~\ref{S-fig:unclustered_msd_shortlag} of the Supporting Information
  for example). This is not a major concern for predicting macroscopic properties
  since we are primarily interested in reproducing the MSDs on much longer 
  timescales.
